---
title: "Mote_subset"
author: "Allyson DeMerlis"
date: "1/6/2021"
output: html_document
---
Install and Load Packages
```{r}
library(tidyverse)
library(DESeq2)
library(ggrepel)
library(apeglm)
library(EnhancedVolcano)
library(ComplexHeatmap)
library(pheatmap)
library(genefilter)
library(WGCNA)
library(flashClust)
library(edgeR) #counts per million
```

Import counts data for orbicella faveolata from feature counts (removed low read depth samples and kept only Mote samples)
```{r}
countdata <- read.delim("ofav_counts", row.names = 1, skip = 1)

# Project Species Status n
# Mote	Ofav	control	  19	
# Mote	Ofav	exposed	  19	
# Smithsonian	Ofav	control	3	
# Smithsonian	Ofav	exposed	6	

#reformatting the counts file for use
colnames(countdata) <- gsub("X.scratch.projects.transcriptomics.mikeconnelly.projects.sctld_jamboree.tagseq.outputs.alignments.Ofav_na_", "", colnames(countdata))
colnames(countdata) <- gsub("_na_Aligned.sortedByCoord.out.uniq.bam", "", colnames(countdata))
colnames(countdata) <- gsub("_", "", colnames(countdata))
colnames(countdata) <- gsub("na", "", colnames(countdata))

countdata %>%select(!Chr:Length) -> countdata
```

Subset Mote samples to remove 5 diseased samples that were clustering with controls
```{r}
countdata %>% 
  select(-K76D1, -K80D1, -K82D1, -K78D1, -K74D1) -> subset_Mote

write_csv(subset_Mote, "subset_mote_ofav_rawcounts.csv")
```

create metadata file
```{r}
#making the metadata file for the sample and treatment information
meta2=data.frame(sample = colnames(subset_Mote),
                condition = stringr::str_detect(pattern = "control",string = colnames(subset_Mote))
                )
#changing TRUE and FALSE for condition to Control and Wounded
meta2$condition[str_detect(meta2$condition,"TRUE")] <- "Control"
meta2$condition[str_detect(meta2$condition,"FALSE")] <- "Diseased"

meta2<-meta2 %>% column_to_rownames("sample")

ncol(subset_Mote) == nrow(meta2)

countdata_subset_matrix <- as.matrix(subset_Mote)
```

Create DESeq Object

1. all mote samples (samples = x, genes = 30175)
2. filtered mote samples (samples = y, genes = 30175)
3. 

```{r}
dds_ofav <- DESeqDataSetFromMatrix(countData=countdata_subset_matrix, colData=meta2, design= ~condition)
summary(results(dds_ofav))

nrow(countdata_subset_matrix)

#different ways to pre-filter low counts
keep <- rowSums(counts(dds_ofav)) >= 10 #removing anything with counts less than 10, suggested by DESeq2 vignette by Michael Love
dds_ofav <- dds_ofav[keep,]
#23,837 genes with counts over 10
summary(results(dds_ofav), alpha = 0.05) #low counts 4479

keep2<- rowSums(counts(dds_ofav)>=4) >=5
dds_ofav2 <- dds_ofav[keep2,] #21003 genes
dds_ofav2 <- DESeq(dds_ofav2)
summary(results(dds_ofav2), alpha = 0.05) #low counts 408
res_ofav2 <- results(dds_ofav2, alpha = 0.05)

# filtering less than 4 samples with counts less than or equal to 5

nrow(dds_ofav)
nrow(dds_ofav2)
nrow(dds_ofav3)

keep3 <- rowSums(cpm(dds_ofav) >= 1) >=3
dds_ofav3<-dds_ofav[keep3,] #23259 genes
dds_ofav3<-DESeq(dds_ofav3)
summary(results(dds_ofav3), alpha = 0.05) #low counts 2255

keep_90 <- rowSums(counts(dds_ofav) >= 10) > (nrow(meta2)*0.9)
dds_ofav_wgcna <- dds_ofav[keep_90, ]
table(keep_90)
# 4,353 genes with counts over 10 in 90% of samples - useful for WGCNA because we need to subset them... can't use all 23,000+ genes

#Perform DESeq2 analysis
dds_ofav <- DESeq(dds_ofav)

resultsNames(dds_ofav)

lfc_shrink_ofav_res <- lfcShrink(dds_ofav, coef="condition_Diseased_vs_Control", type="apeglm") 
#Shrinkage of effect size (LFC estimates) is useful for visualization and ranking of genes
#apeglm method for effect size shrinkage (Zhu, Ibrahim, and Love 2018), which improves on the previous estimator

res_ofav <- results(dds_ofav, alpha = 0.05) #alpha is set for FDR adjusted p-values (that is default of DESeq2)
summary(results(dds_ofav))

sum(res_ofav$padj < 0.05, na.rm=TRUE) #4823 genes

summary(res_ofav)

normalized_counts_ofav_Mote <- counts(dds_ofav, normalized=TRUE)

write.table(normalized_counts_ofav_Mote, file="normalized_counts_mote_ofav.txt", sep="\t", quote=F, col.names=NA)

ofav_results_table <- (as.data.frame(res_ofav2)) 

ofav_results_table %>% rownames_to_column("Locus") -> ofav_results_table

dds_vst_ofav <- vst(dds_ofav, blind = T) #use blind = TRUE to not account for experimental design

plotPCA(dds_vst_ofav)
```

PCA Axes 1 and 2 with GGplot
```{r}
pca12 <- plotPCA(dds_vst_ofav,intgroup=c("condition"),returnData = TRUE)
ggplot(pca12, aes(PC1, PC2,color=condition)) + 
  geom_point(size=3) +  xlab(paste0("PC1 45% variance")) + 
  ylab(paste0("PC2 13% variance")) + 
  theme(legend.position="right")  + 
  theme(text = element_text(size=10))  + 
  theme(legend.key.size = unit(0.5, "cm")) + 
  geom_point(size = 3) +
  theme_classic() + 
  stat_ellipse(aes(PC1, PC2, group=condition), type = "norm") +
  scale_color_manual(values = (c("#7CBAF5", "#AD161A")))
ggsave("PCA_ofav_Mote_removeddiseased.png")

```

Gene features to results table
```{r}
ofav_prot <- read.csv("Ofav_Mote/ofav_proteins_13173_311351.csv")

eggNOGcols <- c("Query", "seed_eggNOG_ortholog",	"seed_ortholog_evalue",	"seed_ortholog_score",	"best_tax_level",	"Preferred_name",	"GOs",	"EC",	"KEGG_ko",	"KEGG_Pathway",	"KEGG_Module",	"KEGG_Reaction",	"KEGG_rclass",	"BRITE",	"KEGG_TC",	"CAZy",	"BiGG_Reaction", "annot lvl",	"matching OGs",	"Best OG",	"COG_cat",	"description")

ofav_eggnog <- read.delim("Ofav_Mote/ofav.protein.emapper.annotations.csv", comment.char  = "#", header = F, col.names = eggNOGcols)

ofav_eggnog %>% dplyr::rename(Protein.product = Query) -> ofav_eggnog

#I THINK THESE ARE JOINING WRONG BECAUSE RESULTS TABLE HAS 23000 genes and then joining it with the protein one leads to 30,000 genes.... i don't think we want that?
#JK yes we do want that.... a gene can make more than one protein.
ofav_annot_DEG <- left_join(ofav_results_table, ofav_prot, by="Locus")
write.csv(ofav_annot_DEG, "ofav_annotated_DEG_Mote.csv")

ofav_annot_DEG$Locus %>% unique %>% length() #there are 21113 unique loci/genes 

ofav_results_table$Locus %>% unique %>% length() #there are 21113 unique loci/genes

#check these tables before uploading them to basecamp
ofav_annot_DEG_emap <- left_join(ofav_annot_DEG, ofav_eggnog, by="Protein.product")

ofav_annot_DEG_emap$Locus %>% unique %>% length() #checking that all 23837 genes were included, yes they were

write.csv(ofav_annot_DEG_emap, "ofav_annotated_DEG_Mote_eggnog.csv")
```


Principal Coordinates Analysis - 
to assess overall similarity between samples: Which samples are similar to each other, which are different? Does this fit to the expectation from the experiment’s design?
```{r}
# Calculate distances among samples
sampleDists <- dist(t(assay(dds_vst_ofav)), method = "manhattan")
sampleDistMatrix <- as.matrix(sampleDists)

# Calculate MDS
mds <- as.data.frame(colData(dds_vst_ofav)) %>% 
  cbind(cmdscale(sampleDistMatrix))

# Calculate MDS and use eigenvectors to determine proport
mds_eig <- cmdscale(sampleDistMatrix, eig = TRUE)
mds_eigenvectors <- data.frame(mds_eig$eig) %>% 
  mutate(prop_var = mds_eig.eig / sum(mds_eig.eig))

# Calculate Treatment centroids for plotting
mds_trmt <- mds %>%
  group_by(condition) %>%
  dplyr::summarise(c1 = mean(`1`), c2 = mean(`2`)) %>%    
  full_join(mds)

# Plot with spiders
ggplot(mds_trmt, aes(fill =condition)) +
  #stat_ellipse(aes(x = `1`, y = `2`, color = condition, fill = condition), geom = "polygon", type = "norm", alpha = 0.0) + 
  # sample-centroid spiders paths
  geom_segment(mapping = aes(x = `1`, y = `2`, xend = c1, yend = c2),
               lwd = 0.25, col = "dark grey") +
  # treatment centroid points
  geom_point(size = 3, aes(x = c1, y = c2, color = condition), fill = "black", shape = 21, stroke = 2, show.legend = TRUE) +
  # sample points
  geom_point(size = 3, aes(x = `1`, y = `2`, color = condition), stroke = 0.5, show.legend = FALSE) +
  scale_color_manual(values = (c("#7CBAF5", "#AD161A"))) +
  theme_bw() +
  labs(x = "MDS1", y = "MDS2")

```

MA plots
```{r}
plotMA(res_ofav, ylim = c(-10,20))
```


Volcano plots with DEGs

```{r}


EnhancedVolcano(res_ofav2,
    lab = NA,
    x = 'log2FoldChange',
    y = 'pvalue',
    col=c('grey', 'black', 'blue', 'purple'))
ggsave("volcanoplot_ofav.png", height = 12, width = 10)
#The default cut-off for log2FC is >|2|; the default cut-off for P value is 10e-6. 

# Make a basic volcano plot
with(res_ofav, plot(log2FoldChange, -log10(pvalue), pch=20, main="Volcano plot", xlim=c(-6,12)))

# Add colored points: light blue if padj<0.01, purple if log2FC>1 and padj<0.05) Adjusted p-value based on FDR, not Bonferroni b/c of DESeq2
with(subset(res_ofav, padj<.05 ), points(log2FoldChange, -log10(pvalue), pch=20, col="lightblue"))
with(subset(res_ofav, padj<.05 & abs(log2FoldChange)>2), points(log2FoldChange, -log10(pvalue), pch=20, col="purple"))
```

heatmap
```{r custom annot file for bingo and cytoscape}
ofav_annot_DEG_emap %>% dplyr::select(Protein.product, GOs) %>%
  mutate_all(na_if,"") %>% 
  drop_na() %>% 
  tidyr::separate_rows(GOs, sep = ",") %>%
  mutate(GOs = str_remove(GOs, "GO:")) %>% 
  unite(x, c(Protein.product, GOs), sep = " = ", remove = T) -> caf_go

View(caf_go)

write.table(caf_go,
            file = "~/Desktop/cyto_ofav_input.txt",
            quote=F, 
            row.names=F, 
            col.names=F)
```


```{r}
ofav_annot_DEG %>% 
  filter(padj < 0.01) %>%
  select(Protein.product) %>% 
  drop_na() %>% 
  column_to_rownames(var = "Protein.product") %>% 
  row.names(.) -> ofav_01

ofav_annot_DEG %>% 
  filter(padj < 0.05) %>%
  select(Protein.product) %>% 
  drop_na() %>% 
  column_to_rownames(var = "Protein.product") %>% 
  row.names(.) -> ofav_05

```



```{r}
topVarGenes <- head(order(-rowVars(assay(dds_vst_ofav))),20)
mat <- assay(dds_vst_ofav)[ topVarGenes, ]
mat <- mat - rowMeans(mat)
df <- as.data.frame(colData(dds_vst_ofav))
df %>% select(condition) -> df

png("ofav_heatmap.png")
pheatmap(mat, annotation_col=df) 

View(order(-rowVars(assay(dds_vst_ofav))))

View(rowVars(assay(dds_vst_ofav)))
```


WGCNA
```{r}
# Normalize expression data for WGCNA using VST tranformation
vsd_ofav_wgcna <- vst(dds_ofav_wgcna, blind = TRUE)

allowWGCNAThreads(nThreads = 4)

# obtain vst-transformed counts
countdata_vst_ofav <- assay(vsd_ofav_wgcna)

# Convert to matrix and transpose, check dimensions
datExpr_ofav <- t(countdata_vst_ofav)
dim(datExpr_ofav)
#21 samples, 4,353 genes...

# Find correlation power R^N that satisfies scale free critereon (SFT.R.sq>0.9)
sft_ofav <- pickSoftThreshold(datExpr_ofav, verbose=5)
sft_ofav$powerEstimate
# Power estimate: 16

# This follows the tutorial: 
# https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/FemaleLiver-02-networkConstr-man.pdf
# 
#From Mike Connelly:
# I have chosen the following network construction parameters for the following reasons:
# First, following the recommendations of the WGCNA developers (https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/faq.html), a signed network was chosen to be able to detect positive and negative gene correlations, and the biweight midcorrelation was used since it is more robust to outliers. 

adjacency <- adjacency(datExpr_ofav,
      # Network construction arguments:  correlation, adjacency function,  and topological overlap map options
                       corFnc = "bicor", # bimidweight correlation
                       power = sft_ofav$powerEstimate, # 20 for coral, 5 for symbiont
                       type = "signed") # signed
# 
TOM <- TOMsimilarity(adjacency,
                     TOMType = "signed",
                     verbose = 5)
dissTOM <- 1-TOM
# 
rm(adjacency) # may need to delete adjacency, TOM to clear up vector memory


geneTree_ofav <- flashClust(as.dist(dissTOM), method = "average")
plot(geneTree_ofav, labels = FALSE, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity")

# minModuleSize = 30 because Mike prefers large modules that lend themselves to enrichment tests with GO_MWU for easier interpretation.
minModuleSize <- 30
# Module identification using dynamic tree cut, with cutHeight = 0.99 and deepSplit = 2 based on the recommendations of the WGCNA developers
dynamicMods <- cutreeDynamic(dendro = geneTree_ofav,
                             distM = dissTOM,
                             cutHeight = 0.99,
                             deepSplit = 2,
                             pamRespectsDendro = FALSE,
                             minClusterSize = minModuleSize)
table(dynamicMods)
# Convert numeric labels into colors
dynamicColors <- labels2colors(dynamicMods, colorSeq = standardColors())

# Calculate eigengenes
MEList <- moduleEigengenes(datExpr_ofav, colors = dynamicColors)
MEs <- MEList$eigengenes
# Calculate dissimilarity of module eigengenes
MEDiss <- 1-cor(MEs);
# Cluster module eigengenes
METree <- flashClust(as.dist(MEDiss), method = "average");
# Plot the result
plot(METree, main = "Clustering of module eigengenes",
xlab = "", sub = "")
MEDissThres = 0.15
# Plot the cut line into the dendrogram
abline(h=MEDissThres, col = "red")

# Call an automatic merging function
# merge cutHeight = 0.3 because I want to group together modules with >85% similar module eigengene expression
mergedMods <- mergeCloseModules(datExpr_ofav, dynamicColors, cutHeight = 0.15, verbose = 5)
# The merged module colors
mergedColors <- mergedMods$colors
table(mergedColors) 
```

```{r}
png(file = "WGCNA_dendrogram_Ofav.png")

# Plot the dendrogram and the module colors underneath
plotDendroAndColors(dendro = geneTree_ofav, 
                    colors = cbind(dynamicColors, mergedColors),
                    groupLabels = c("Dynamic modules", "Merged dynamic"),
                    dendroLabels = FALSE, hang = 0.03,
                    addGuide = TRUE, guideHang = 0.05,
                    abHeight = c(0.99))
```
```{r merged_eigengenes}
MEList <- moduleEigengenes(datExpr_ofav, colors = mergedColors)
mergedMEs <- MEList$eigengenes
```
```{r choose_modules}
moduleColors <- mergedColors#netColors#
MEs1 <- mergedMEs#netMEs#
MEs1 <- MEs1 %>% select(-MEgrey)
MEmodule_colors <- colnames(MEs1)
module_colors <- gsub("ME", "", colnames(MEs1))
colnames(MEs1) <- module_colors
```
```{r ME_clustering}
datME <- MEs1
dissimME <- (1-t(cor(datME, method="p")))/2
hclustME <- flashClust(as.dist(dissimME), method="average" )
# Plot the eigengene dendrogram
par(mfrow=c(1,1))
plot(hclustME, main="Clustering tree based on the module eigengenes")
```
```{r uniq_modules}
# Extract all unique modules
uniqModules <- unique(colnames(MEs1))
# set in order according to ME clustering dendogram
hclustME$order
uniqModules <- uniqModules[hclustME$order]
#
uniqModules
# create ordered factor for downstream analysis
modules_factor <- factor(levels = uniqModules[hclustME$order], ordered = TRUE)
```

## Module genes and hub genes
### *O. faveolata*
```{r modules_genes}
# Generate GeneIDs
Genes <- colnames(datExpr_ofav)
# Output genes and annotations for each module
for (module in uniqModules)
{
# Select module genes
inModule <- (moduleColors == module)
# Get gene IDs
modGeneIDs <- Genes[inModule] #this is the correct set of gene IDs!
# Write gene annotations into a file
fileName = paste(module, ".csv", sep="");
module_annotation <- ofav_annot_DEG[ofav_annot_DEG$Locus %in% modGeneIDs, ]
write_csv(module_annotation, path = fileName)
}
```


```{r hub_genes}
hubs <- chooseTopHubInEachModule(datExpr_ofav,
                         mergedColors,
                         corFnc = "bicor", # bimidweight correlation
                         power=sft_ofav$powerEstimate, # 16
                         # power = 10,
                         type = "signed")
hub_genes <- ofav_annot_DEG[ofav_annot_DEG$Locus %in% hubs, ]

hub_genes %>% arrange(Locus) %>% mutate(module =  c("darkgreen", "darkgreen", "midnightblue","lightgreen","blue","royalblue","royalblue","brown", "lightyellow", "magenta", "darkred", "pink","greenyellow","turquoise","tan","salmon","lightcyan","grey60","grey60","yellow","black","cyan","cyan")) -> hub_genes

row.names(hub_genes) <- NULL 
hub_genes <- hub_genes %>% dplyr::select(module, everything())
write_csv(hub_genes, "hub_genes_WGCNA_ofav.csv")
```

## Module eigengene correlations
```{r}
# Tidy sample metadata into a format WGCNA can use for correlations
ofav_mote_traits <- metadata %>% 
  select(Working.Label, Project, Species, percent_disease_progress, Correct_treatment) %>% 
  filter(Project == "Mote") %>% 
  filter(Species == "Ofav") %>% 
  filter(Working.Label!="K74", Working.Label!="K76", Working.Label!="K78",Working.Label!="K80",Working.Label!="K82")
#-K76D1, -K80D1, -K82D1, -K78D1, -K74D1
ofav_mote_traits$percent_disease_progress[ofav_mote_traits$percent_disease_progress=="control"] <- 0
for (i in 1:nrow(ofav_mote_traits)){
if (ofav_mote_traits$Correct_treatment[i] == "control") {
  ofav_mote_traits$control[i] <- "1"
} else {
  ofav_mote_traits$control[i] <- "0"
}
}
# 
for (i in 1:nrow(ofav_mote_traits)){
if (ofav_mote_traits$Correct_treatment[i] == "diseased") {
  ofav_mote_traits$disease[i] <- "1"
} else {
  ofav_mote_traits$disease[i] <- "0"
}
}
ofav_mote_traits_final <- ofav_mote_traits %>%
  column_to_rownames("Working.Label") %>% 
  select(-Correct_treatment)
```


```{r}
MET = orderMEs(cbind(MEs, condition))
# Plot the relationships among the eigengenes and the trait
sizeGrWindow(5,7.5);
par(cex = 0.9)
plotEigengeneNetworks(MET, "", marDendro = c(0,4,1,2), marHeatmap = c(3,4,1,2), cex.lab = 0.8, xLabelsAngle
= 90)
```



